Variable Density Tesseroids: Gravity fields calculation in spherical coordinates using variable densities in depth

### 2017-10-24 manuscript/manuscript.tex
Changes in manuscript

These forward gravity models have been developed for two or three dimensional bodies in cartesian coordinates that properly fit local scales applications.
Nevertheless, the latest satellite missions have provided us gravity measurements with global coverage, which allows geologists and geophysics to perform modelling and interpretation on regional scales.
This raises the importance of building forward models that reproduce the gravity anomalies for such scales.

The main issue that should be overcome is taking into account the curvature of the Earth, thus the forward model must be defined in spherical coordinates.
A common way to achieve this is to discretize the Earth in spherical prisms known as tesseroids, which are defined by pairs of latitude, longitude and radius boundaries (see Figure \ref{fig:tesseroid-uieda}).
Calculating the gravity fields generated by an arbitrary tesseroid on any external point involves the resolution of volume integrals that are generally approximated by numerical computations.
The literature offers two approaches: one involves Taylor series expansion \citep{Heck2007, Grombein2013} while the other makes use of Gauss-Legendre Quadrature (GLQ) \citep{Asgharzadeh2007, Uieda2016, Uieda2017}.
The later consists in approximating the integral by a weighted sum of the effect of point masses located at scaled nodes of the Legendre polynomials.

@@

In order to develop a forward model that computes the gravity field generated by any tesseroid with arbitrary variable density, the Taylor series expansion is not well suited, because it would need to derive the series expansion terms for each density function.
On the other hand, the GLQ allows us to calculate them for any density function without any other information but the function itself.

@@

We have enhanced the implementation of \citet{Uieda2016} with a new forward gravity model that allows us to compute the gravity fields generated by any tesseroid with an arbitrary density function on any external point.
It is based on the GLQ approximation, makes use of the modified distance-size discretization and the new density-based discretization algorithms that guarantee the accuracy even for high varying density functions.
It is written in Python and Cython languages, constituting an easy to use library that runs precompiled code for the more time consuming functions.

@@

This new forward gravity model makes use of some previously existing classes and functions from Fatiando a Terra, what saved us time and allowed us to focus on the new enhancements.
Furthermore, we have the intention to include it into a future release of the Fatiando a Terra library in order to make its distribution and maintenance easier.

@@

The spherical prisms known as tesseroids are mass elements defined in a geocentric spherical coordinate system bounded by a pair of parallels, a pair of  meridians, and two concentric spherical surfaces.
We define an external computation point $P(r, \phi, \lambda)$ where the gravity fields generated by the tesseroid are going to be calculated with respect to the local north oriented Cartesian coordinate system located at $P$.
In the special case of homogeneous density, the gravity potential, its gradient and the Marussi tensor components can be calculated through the integrals obtained by \citet{Grombein2013} \citep[see also][]{Uieda2016}.

@@

\noindent $\alpha, \beta \in \{x, y, z\}$,
$\rho(r')$ is the density function that depends on the radius coordinate,
$\delta_{\alpha\beta}$ is Kronecker's delta,
$G = 6.674\times10^{-11}\, \text{m$^3$kg$^{-1}$s$^{-1}$}$ is the gravitational constant and

\begin{equation}
    \Delta_x = r'[\cos\phi\sin\phi' - \sin\phi\cos\phi'
               \cos(\lambda' - \lambda)],
\end{equation}
\begin{equation}
    \Delta_y = r' \cos \phi' \sin(\lambda' - \lambda),
\end{equation}
\begin{equation}
    \Delta_z = r' \cos \psi - r,
\end{equation}
\begin{equation}
    \kappa = {r'}^2 \cos \phi',
\end{equation}
\begin{equation}
    \ell = \sqrt{{r'}^2 + r^2 - 2 r' r \cos \psi},
\label{eq:ell}
\end{equation}
\begin{equation}
    \cos\psi = \sin\phi\sin\phi' + \cos\phi\cos\phi'
                 \cos(\lambda' - \lambda).
\label{eq:cospsi}
\end{equation}

@@

\noindent where $x_i^s$ is the scaled $x_i$ root in the $[a,b]$ interval, also called nodes of the quadrature:

@@

For example, if we use a second order GLQ, the roots of the $P_2(x)$ Legendre polynomial and its corresponding weights are $x_i = \pm 0.577350269$ and $W_i = 1$.

@@

From equation \ref{eq:glq-var-dens} we can observe that the GLQ approximates the gravity fields of a tesseroid with variable density as the ones generated by point masses located at the nodes of the Legendre polynomials.
Furthermore, we can also conclude that the information of the density function is summarised as the values it assumes on those nodes.
Although this may sound discouraging, we intent to prove otherwise and show that this method, along with well fitted discretization algorithms, approximates the gravity fields with good accuracy.\todo{no se si poner esta ultima oracion}

@@

\begin{equation}
    \cos\psi = \sin\phi', \quad
    \ell = \sqrt{r'^2 - 2 r r' \sin\phi' + r^2}.
\end{equation}

@@

\begin{equation}
    V_\text{sh}(r) = 2\pi G 
    \int\limits_{-\frac{\pi}{2}}^\frac{\pi}{2}
    \int\limits_{R_1}^{R_2}
    \frac{\rho(r') {r'}^2 \cos\phi'}{\sqrt{r'^2 - 2 r r' \sin\phi' + r^2}}
    \, dr' d\phi',
\end{equation}

\noindent while the integration in $\phi'$ can be performed independently of the density function.
Using \citet{sagemath} we obtained the following expression:

@@

If the density of the spherical shell is a linear function on the radius coordinate, i.e.:

@@

If, on the other hand, the spherical shell has an exponential density in the radius coordinate,

@@

In order to overcome this, we developed a complementary discretization algorithm that takes into account the variation of the density inside the tesseroid.
Basically, it subdivides any tesseroid on the radius coordinate on the depth at which the \emph{maximum density variation} takes place.

>>>>

These forward gravity models have been developed for two or three dimensional bodies in cartesian coordinates that properly fit local scales applications.
Nevertheless, the latest satellite missions have provided us gravity measurements with global coverage, which allow geologists and geophysics to perform modelling and interpretation on regional scales.
This raises the importance of building forward models that reproduce the gravity anomalies for such scales.

The main issue that should be overcome is taking into account the curvature of the Earth, thus the forward model must be defined in spherical coordinates.
A common way to achieve this is to discretize the Earth in spherical prisms known as tesseroids, which are defined by pairs of latitude, longitude and radius boundaries (see Figure \ref{fig:tesseroid-uieda}).
Calculating the gravity fields generated by an arbitrary tesseroid on any external point involves the resolution of volume integrals that are generally approximated by numerical computations.
The literature offers two approaches: one involves Taylor series expansion \citep{Heck2007, Grombein2013} while the other makes use of Gauss-Legendre Quadrature (GLQ) \citep{Asgharzadeh2007, Uieda2016, Uieda2017}.

In order to develop a forward model that computes the gravity field generated by any tesseroid with arbitrary variable density, the Taylor series expansion is not well suited, because it would need to derive the series expansion terms for each density function.
On the other hand, the GLQ consists in approximating the integral by a weighted sum of the effect of point masses located at scaled nodes of the Legendre polynomials, where the density information is gathered into these weights.
Thus, this kind of numerical approximation allows to calculate the gravity fields for any density function without any other information but the function itself.

@@

@@

We have enhanced the implementation of \citet{Uieda2016} with a new forward gravity model that allows us to compute the gravity fields generated by any tesseroid with an arbitrary density function on any external point.
It is based on the GLQ approximation, makes use of the modified distance-size discretization and the new density-based discretization algorithms that guarantee the accuracy even for high varying density functions.
It is written in Python and Cython languages, constituting an easy to use library that runs precompiled code for the more time consuming functions.
It also makes use of some previously existing classes and functions from Fatiando a Terra, what saved us time and allowed us to focus on the new enhancements.
Furthermore, we have the intention to include it into a future release of the Fatiando a Terra library in order to make its distribution and maintenance easier.

@@

@@

The spherical prisms known as tesseroids are mass elements defined in a geocentric spherical coordinate system bounded by a pair of parallels, a pair of  meridians, and two concentric spherical surfaces.
We define an external computation point $P(r, \phi, \lambda)$ where the gravity fields generated by the tesseroid are going to be calculated with respect to the local north oriented Cartesian coordinate system located at $P$.
In the special case of homogeneous density, the gravity potential, its gradient and the Marussi tensor components can be calculated through the integrals obtained by \citet{Grombein2013} \citep[for same notation as the one we will use, see][]{Uieda2016}.

@@

\noindent $\alpha, \beta \in \{x, y, z\}$,
$\rho(r')$ is the density function that depends on the radial coordinate,
$\delta_{\alpha\beta}$ is Kronecker's delta,
$G = 6.674\times10^{-11}\, \text{m$^3$kg$^{-1}$s$^{-1}$}$ is the gravitational constant and

\begin{equation}
    \Delta_x = r'[\cos\phi\sin\phi' - \sin\phi\cos\phi'
               \cos(\lambda' - \lambda)],
\end{equation}
\begin{equation}
    \Delta_y = r' \cos \phi' \sin(\lambda' - \lambda),
\end{equation}
\begin{equation}
    \Delta_z = r' \cos \psi - r,
\end{equation}
\begin{equation}
    \kappa = {r'}^2 \cos \phi',
\end{equation}
\begin{equation}
    \ell = \sqrt{{r'}^2 + r^2 - 2 r r' \cos \psi},
\label{eq:ell}
\end{equation}
\begin{equation}
    \cos\psi = \sin\phi\sin\phi' + \cos\phi\cos\phi'
                 \cos(\lambda' - \lambda).
\label{eq:cospsi}
\end{equation}

@@

\noindent where $x_i^s$ is the scaled $x_i$ root into the $[a,b]$ interval, also called nodes of the quadrature:

@@

\noindent For example, if we use a second order GLQ, the roots of the $P_2(x)$ Legendre polynomial and its corresponding weights are $x_i = \pm 0.577350269$ and $W_i = 1$.

@@

In the same way as for an homogeneous density, we can observe from equation \ref{eq:glq-var-dens} that the GLQ approximates the gravity fields of a tesseroid as the ones generated by point masses located at the nodes of the Legendre polynomials.
Nevertheless, in the variable density case, the information of the density function is summarised into the weights of the sum by taking into account only the values it assumes on those nodes.
Although this may sound discouraging, we intent to prove otherwise and show that this method, along with well fitted discretization algorithms, approximates the gravity fields with good accuracy.\todo{no se si poner esta ultima oracion}

@@

\begin{equation}
    \cos\psi = \sin\phi', \quad
    \ell = \sqrt{r'^2 + r^2 - 2 r r' \sin\phi'}.
\end{equation}

@@

\begin{equation}
    V_\text{sh}(r) = 2\pi G 
    \int\limits_{-\frac{\pi}{2}}^\frac{\pi}{2}
    \int\limits_{R_1}^{R_2}
    \frac{\rho(r') {r'}^2 \cos\phi'}{\sqrt{r'^2 + r^2 - 2 r r' \sin\phi'}}
    \, dr' d\phi',
\end{equation}

\noindent while the integration in $\phi'$ can be performed independently of the density function.
Taking into account the following integral:

\begin{equation}
    \int \frac{\cos \phi}{\sqrt{a - b \sin \phi}} \, d\phi =
    -\frac{2}{b} \sqrt{a - b \sin \phi} + C,
\end{equation}

@@

If the density of the spherical shell is a linear function on the radial coordinate, i.e.:

@@

If, on the other hand, the spherical shell has an exponential density in the radial coordinate,

@@

In order to overcome this, we developed a complementary discretization algorithm that takes into account the variation of the density inside the tesseroid.
Basically, it subdivides any tesseroid on the radial direction on the depth at which the \emph{maximum density variation} takes place.
